Parallelizing Numerical Methods for Solving ODEs: The Parareal Algorithm

### 2014-12-09 poster/main.tex
POSTER COMPLETE
Conflicts:
	paper.out
	paper.tex

\title{The Parareal Algorithm for Parallel Solution of ODEs} % Poster title

@@

We implement the Parareal algorithm to solve various ODEs and compare in
terms of accuracy, speedup and efficiency. We explore the effect of using
different solution operators (a coarse and a fine as required by the method).
Our code is written in python using mpi4py to parallelize. We observe
light speedups in our small scale tests up to 64 processors on Harvard's
Odyssey cluster. We analyze possible reasons as to why our implementation
may not be ideal and some tradeoffs that are taken in the Parareal
implementation. Further optimizations are proposed and rationalized as next
steps including ports of to C++ and BLAS libraries which is a more
performance-oriented, lower-level language than Python. Data is shown for
solving for various complexities of ODEs from the basic exponential to
modeling sound wave propagation as we did in homework 4.

@@

The Parareal algorithm, developed by Lions, Maday and Turinici in 2001, is
a general algorithm that allows parallelization in time.sIt does so by using a
cheaper (lower order, lesser resolution) approximation first, and then makes
corrections in parallel. The entire process is then iterative and tuned for a
given amount of iterations. Therefore, the algorithm is $k$ repetitions of a
serial coarse method updated by a finer method run in parallel for subsections.

Subscript: time index. Superscript: iteration index.

\begin{itemize}
    \item \textbf{Input}: Temporal discretization $t_n = t_0 + n \Delta t, \, n =
        1,2,\ldots,N$
    \item \textbf{Input}: Coarse scheme $g_{\Delta t}$
    \item \textbf{Input}: Finer scheme $g_{\textnormal{fine}}$
\end{itemize}

\begin{enumerate}
    \item Compute $u^1_{n+1} = g_{\Delta t}(t_n, u^1_n)$
    \item Compute the corrections $\delta g_n(u^1_n) =
        g_{\textnormal{fine}}(t_n, u^1_n) - g_{\Delta t}(t_n, u^1_n)$ in parallel
    \item Add the prediction and correction terms as $u^2_{n+1} = g_{\Delta
        t}(t_n, u^2_n) + \delta g_n(u^1_n)$
    \item Repeat steps 2 and 3, incrementing the iteration label and using
        $u^{k+1}_0 = u^1_0$ as the initial condition
\end{enumerate}
\end{block}

@@

%----------------------------------------------------------------------------------------
%	MATERIALS
%----------------------------------------------------------------------------------------

@@

%----------------------------------------------------------------------------------------
%	MATHEMATICAL SECTION
%----------------------------------------------------------------------------------------

\begin{block}{Theoretical Results}

@@

\textbf{Runtime and Speedup}

@@

In order for there to be a speedup relative to the fine operator, we require
that $t + k(t + Qt/N) < Qt$, or $k < \frac{Q - 1}{1 + Q/N}$ or $N > \frac{Qk}{Q
- 1 - k}$.

@@

\begin{block}{Preliminary Results}

@@

%----------------------------------------------------------------------------------------
%	CONCLUSION
%----------------------------------------------------------------------------------------

\begin{block}{Conclusion}

@@

%----------------------------------------------------------------------------------------
%	ADDITIONAL INFORMATION
%----------------------------------------------------------------------------------------

\begin{block}{Additional Information}

\begin{block}{References}

\nocite{*} % Insert publications even if they are not cited in the poster
\small{\bibliographystyle{unsrt}
\bibliography{sample}\vspace{0.75in}}

%----------------------------------------------------------------------------------------
%	ACKNOWLEDGEMENTS
%----------------------------------------------------------------------------------------

\setbeamercolor{block title}{fg=red,bg=white} % Change the block title color

\begin{block}{Acknowledgements}

\small{\rmfamily{Nam mollis tristique neque eu luctus. Suspendisse rutrum congue nisi sed convallis. Aenean id neque dolor. Pellentesque habitant morbi tristique senectus et netus et malesuada fames ac turpis egestas.}} \\

%----------------------------------------------------------------------------------------
%	CONTACT INFORMATION
%----------------------------------------------------------------------------------------

\begin{alertblock}{Contact Information}

\begin{itemize}
\item Web: \href{http://www.university.edu/smithlab}{http://www.university.edu/smithlab}
\item Email: \href{mailto:john@smith.com}{john@smith.com}
\item Phone: +1 (000) 111 1111
\end{itemize}

\begin{center}
\includegraphics[width=0.2\linewidth]{seas-rgb.png}
\end{center}

>>>>

\title{Parallelizing Numerical Methods for Solving ODEs: The Parareal Algorithm} % Poster title

@@

We look to achieve strong scaling in parallelizing numerical ODE methods. We test both parallelism by result space as well as parallelism by time, with the Parareal algorithm, to solve various ODEs.  We measure performance and compare in
terms of accuracy, speedup and efficiency. We provide some theoretical analysis for the Parareal algorithm and compare its tradeoff to paralleism by space, For the Parareal agorithm, we observe
light speedups in our small scale tests up to 64 processors on Harvard's
Odyssey cluster. We analyze possible reasons as to why our implementation
may not be ideal and some tradeoffs that are taken in the Parareal
implementation. Further optimizations are proposed and rationalized as next
steps including ports of to C++ using BLAS. Our code is written in python using mpi4py.

@@

The Parareal algorithm, developed by Lions, Maday and Turinici in 2001, is
a general algorithm that allows parallelization in time.sIt does so by using a
cheaper (lower order or lesser resolution) approximation first, and then makes
corrections in parallel. The entire process is then iterative and tuned for a
given amount of iterations. Therefore, the algorithm is $k$ repetitions of a
serial coarse method updated by a finer method run in parallel for subsections.

Parareal Algorithm:\\
Subscript: Time [0,n], Superscript: Iteration [1,k]
\begin{itemize}
    \item \textbf{Input}: Temporal discretization $t_n = t_0 + n \Delta t, \, n =
        1,2,\ldots,N$
    \item \textbf{Input}: Coarse scheme $g_{\Delta t}$
    \item \textbf{Input}: Finer scheme $g_{\textnormal{fine}}$
\end{itemize}

\begin{enumerate}
    \item Compute $u^1_{n+1} = g_{\Delta t}(t_n, u^1_n)$
    \item Compute the corrections $\delta g_n(u^1_n) =
        g_{\textnormal{fine}}(t_n, u^1_n) - g_{\Delta t}(t_n, u^1_n)$ in parallel
    \item Add the prediction and correction terms as $u^2_{n+1} = g_{\Delta
        t}(t_n, u^2_n) + \delta g_n(u^1_n)$
    \item Recompute corrections and adjustment, for each k \\ using
        $u^{k+1}_0 = u^1_0$ as the initial condition
\end{enumerate}
\end{block}

@@

@@

%----------------------------------------------------------------------------------------
%	MATHEMATICAL SECTION
%----------------------------------------------------------------------------------------
\begin{block}{Theoretical Analysis}

@@

\textbf{Parallel Performance Tradeoff Analysis}

@@

The first t seconds comes from the first coarse aprpoximation, without which the parareal algorithm degenerates to $g_{\Delta t}$,  In order for there to be a speedup relative to the fine operator, we require
that $t + k(t + Qt/N) < Qt$, or $k < \frac{Q - 1}{1 + Q/N}$ or $N > \frac{Qk}{Q
- 1 - k}$. This can be tuned with by either adjusting the the quality factor or reducing the number of iterations, but the number of iterations, k, is important for error convergence to that of $g_{fine}$.  However, this optimal k is a problem-specific parameter and must be optimized for.

@@

\begin{block}{Preliminary \\ Implementation Results}

@@

%----------------------------------------------------------------------------------------
%	Discussion
%----------------------------------------------------------------------------------------

\begin{block}{Discussion and Next Steps}

Our preliminary results do not look as optimal as we had hoped. We see a significant overhead in the iterative Parareal algorithm when compared to the serial computations of $g_{fine}$ directly.  However, we do see that the Parareal algorithm itself displays both strong and weak scaling, as expected. Yet the biggest tradeoff we look to better demonstrate will be the heavy dependence on the two methods chosen for $g_{\Delta t}$ and $g_{fine}$. \\

Most of the immediate work will involve increased data collection with large N. The Parareal algorithm provides maximum speedup when N approaches the number of discretizations in time for $g_{\Delta t}$ which would allow all corrections to be made at one step. We are also in the process of implementing and switching out various $g_{fine}$ schemes. We are looking to employ higher order methods at the same temporal resolution as opposed to a higher-resolution equal-order method. This should help show our ever convergence of the Parareal method and may also emphasize overall speedup as well.
\end{block}

%----------------------------------------------------------------------------------------
%	Conclusion and Next Steps
%----------------------------------------------------------------------------------------

\begin{alertblock}{Conclusion}

We have implemented and begun to explore some techniques of looking
for strong and weak scaling efficiencies for solving ODEs. The Parareal
algorithm is beautiful in its theoretical advantages---in terms of stability,
error and efficiency. However, the method, as a very generalized method,
requires deeper analysis for the specific problem.

\setbeamercolor{block title}{fg=red,bg=white} % Change the block title color
\begin{block}{References}

\begin{itemize}
\item Lions, J., Maday, Y and Turinici, G. "A 'Parareal' in Time Discretizion of PDE's". 2001. \textit{Comptes Rendus de l'Academie des Sciences Series I Mathematics} 332(7): 661-668\\

\item Field, Scott. "Parareal Methods Field Talk". Dec, 2009. Retried from \url{https://www.sharcnet.ca/help/index.php/Measuring_Parallel_Scaling_Performance}

\end{itemize}

%\nocite{*} % Insert publications even if they are not cited in the poster
%\small{\bibliographystyle{unsrt}
%\bibliography{sample}\vspace{0.75in}}

@@
